surfing

### 2021-05-09 inst/surfing.Rmd
rejigging for journal of sports analytics

(This document is in two parts.  The first is a slightly edited
version of Tim Driver's dissertation on surfing statistics, the second
a short analysis---using new software not available to Tim---of a
dataset he provided).

Abstract: As surfing continuous to grow as a professional sport, there
is a lack of analysis on competition results. This report applies
generalized Bradley-Terry models to get the maximum likelihood of
surfer’s strengths in competitions with powerful waves. By gathering
competition data, analysis can be conducted in R using the `hyper2`
package, resulting in the surfer’s strengths. Using this methodology,
weaker surfers were able to be identified when considering all
powerful wave locations, however, there was not enough data to get a
significant result for single events. Furthermore, it is hoped that a
result forecasting model can be produced using similar techniques.

Competitive surfing involves surfers competing against one or two
other surfers in heats lasting 30-50 minutes. Points are awarded on a
10-point scale, with [0.0- 1.9] a poor score, [2.0-3.9] a fair score,
[4.0 - 5.9] an average score, [6.0 – 7.9] a good score and finally
[8.0 – 10.0] an Excellent score. The highest two waves ridden by an
individual are then combined to produce an individual heat total, the
highest of which then wins the heat [1].

Five judges rate each of the waves ridden by a surfer and award points
based off of a range of criteria including commitment and degree of
difficulty, innovation and progress of manoeuvres, combination of
major manoeuvre, variety of manoeuvres and speed, power and flow
[1]. The top score and bottom score of the judges are removed and the
remaining three judges scores are averaged to give the final score for
the surfer.

Structure of the event is as follows, initially there is a seeding
round consisting of 12 heats with three surfers competing in each
heat. The top two surfers of each heat advance to the round of 32
while the 12 surfers who failed to finish first or second are placed
in an elimination round. The elimination round is again four,
three-man heats with last placed in each heat being eliminated from
the competition. Surfers are then in man on man heats, with the loser
of each heat eliminated from the competition. This format then repeats
with a round of 16, quarter finals, semi-finals and finally the
finals.

The World Surf League (WSL) is the main governing body of surfing
competitions. The WSL conducts a world tour that the best ranked 34
surfers compete in. At each tour venue, two wild cards also enter the
competition. The wildcard surfers are chosen from the main sponsors of
the event and the winner of a local competition that is held before
the main event starts. On occasion, if or more of the 34 surfers are
injured, an injury replacement surfer will be selected by the WSL to
be added to the competition to keep the heat format the same.

The 2019 tour saw competitive events held at eleven different surf
locations. In each of these locations the wave type of wave
differs. Some events are known to be more favourable to those who have
a high number of manoeuvres at their disposal and others are more
suited to surfers who regularly surf bigger waves. Of the 11 locations
in 2019, three competitions stand out as the waves at locations are
known for the size and power. At these locations, judges score riding
inside the wave ("tubed") higher than other turn or air style
manoeuvres. The locations are Pipeline in Hawaii, Teahupo’o in Tahiti
and Margaret River in Western Australia.

At these locations, ocean swell comes from deep water and hits shallow
reefs. The lower part of the wave slows down due to the interference
with the ocean floor while the top part keeps a similar velocity and
eventually the wave breaks. As the top part of the wave is moving
forward at a larger velocity, it breaks closer to the reef, causing
there to be a gap between the top part and the bottom close to the
surface of the water [2].  Surfing in this critical zone is when you
are getting tubed and subsequently scored highly, due to the
commitment and degree of difficulty involved.

Analysis on competitive surfing has not attracted much attention since
surfing became a professional sport 1960s [3]. With the recent
inclusion in the Olympic games, surfing continues to grow in
professionalism. High-level analysis of surfing needs to develop to
allow new insights to be discovered about surfers performance in
competitions to continue this trend of professionalism in the sport.

[discussion of Bradley-Terry elided]

# Literature Review

Currently relevant research into competitive surfing has been limited
to the scoring of ariel manoeuvres compared with other manoeuvres in
events [10] and analysis in the difference in average wave scores for
the top 10 and bottom 10 surfers using a One-way ANOVA method [11].

In the work discussed in the paper "Analysis of Manoeuvres and Scoring
in Competitive Surfing" [10], analysis was done on the waves ridden in
the quarter finals, semi-finals and final heats of the 2012 World
Championship Tour at all locations. On this year of the Tour, Western
Australia was not included and so Margaret River was not surfed. Waves
which involved a tube or arial manoeuvre were shown to score higher
than doing other turn style on the wave across all locations and have
a much lower completion rate than other manoeuvres. This shows the
importance for getting tubed at elite surfing levels to win heats and
thus win competitions. As this research piece is focusing on the Tour
locations that have waves with higher opportunities to get tubed, poor
results here are indicative of poor ability to get tubed.

In the research done in the paper "Scoring Analysis of the Men's 2013
World Championship Tour of Surfing" [11], scores from the top 10
surfers and bottom 10 surfers 2013 World Championship Tour are
analysed to determine if there is any difference in the average wave
score and average total heat score. A one-way ANOVA was performed on
the two groups to determine if there was any major discrepancies. It
was found that the top 10 surfers had a score of 14.7% higher than the
bottom 10 surfers and the standard deviation of the top surfers score
was 9.39% less than that of the bottom 10 surfers. The authors then
took this work to surf coaches to highlight the importance of
performing better as an increase in the average wave score by 10% for
a bottom ranked surfer would greatly improve their overall ranking on
the Tour.

The method of approach in this research report is to use Bradley-Terry
models [4] for paired comparisons to analyse the results of
heats.  Bradley-Terry models are popular for paired comparisons between
competitors.  As earlier heats in the competition have three man heats
a general Bradley Terry model will be used to look at this comparison.

Existing research has been conducted on other sports using
Bradley-Terry models, including tennis [12], baseball [13] and
American football [14].

The existing research conducted by McHale and Morton [12] titled "A
Bradley-Terry type model for forecasting tennis match results" looked
at the top tier results from the Association of Tennis Professionals
(ATP) for 9 seasons between 2000-2008. The data used was the amount of
games in sets won by competitors when playing against other
competitors. Information about the surface that the court was played
on, be it clay, grass or a hardcourt was also included and a form
decaying factor based of previous results. Once

the modified Bradley-Terry models were applied to the data, the model
produced an alternative ranking system. Five of the top 15 official
ATP ranking were not included in the top 15 players returned by the
model. They then go on to use this new ranking to forecast the results
of upcoming tennis matches.

The existing research by Butler and Whelan [14] describes modification
to the original Bradley-Terry model and applies the model to American
Football results in the round robin stages of the 1998 Western
Athletic conference. In the round robin stage, teams compete against
all the teams in their division, along with one team in the other
division. The divisions are separated geographically, and each
contained 8 teams. The model initially identified if any pairs of
teams i and j were strictly better than the other through analysing
previous matchups. If results existed that did not display
transitivity relationship, then Bradley-Terry models were applied to
these subsets in order to determine a superior team.

The original Bradley-Terry model developed only considered paired data
and did not describe a case with more than two competitors. This is
appropriate for the later heats but fails in the earlier three-man
heat competitions. A general model needs to be applied as described by
Luce [7], to consider the case for three or more.

The generalized Bradley-Terry model has been put into R under the
Hyper2 package by Robin K. S. Hankin [15] and has since been used for
a number of projects including analysis of competitive cooking and
rowing [9]. The research done on the rowing piece analysed the men’s
single scull in the 2016 Olympic games. 25 races were run at this
competition with 31 rowers entering, each with a maximum of six rowers
competing per. If a rower won the race, be it a millisecond or a
minute, they were still recorded the same as the only concern of the
model is the placing in whish the rowers finished. The maximum
likelihood estimate for each rower winning the event was produced,
showing an alternative ranking system to the knock out format at the
Olympics. The models ranking had Drysdale first, Synek second and
Martin third. The actual results in the final were Drysdale first,
Martin second and Synek Third.

# Research Question:

The purpose of this research report is to do analysis using
generalized Bradley-Terry models on the 2019 WSL Tour at competition
locations known to have powerful surf. These locations are Margaret
River, Teahupo'o and Pipeline. Generalized Bradley-Terry models will
be used on the surfer's heat results.

Analysis will be done on each competition individually, a combination
of Margaret River and Teahupo'o as well as a combination of all three
locations. Each individual location will be analysed to create an
alternative ranking system to the competition knockout format. All
three locations will be analysed together to determine a ranking
system for the best surfer in powerful waves, based off these
locations having waves of consequence. Margaret River and Teahupo’o
will be analysed together to investigate an early model used for
forecasting results at these locations, specifically can the results
from these two locations be used to forecast the pipeline competition
results.

[1] "Rules and Regulations" [Online]. Available:
https://www.worldsurfleague.com/pages/rules- and-regulations.

[2] M. Longuet-Higgins, "Progress Toward Understanding How Waves
Break," in Twenty-First Symposium on Naval Hydrodynamics, National
Research Council, Naval Studies Board, Division on Engineering and
Physical Sciences, Commission on Physical Sciences, Mathematics, and
Applications, 1997, pp. 7-29.

[3] D. Booth, "Ambiguities in Pleasure and Discipline: The Development
of Competitive Surfing," Journal of Sport HIstory, 1995.

[4] E. M. Terry and A. R. Bradley, "Rank Analysis of Incomplete
Block Designs: I. The Method of Paired Comparisons," Biometrika,
pp. 324-345, 1952.

[5] D. R. Hunter, "MM algorithms for generalized Bradley-Terry
models," The annals of Statistics, pp. 384-406, 2004.

[6] H. Turner and D. Firth, "Bradley-Terry Models in R: The
BradleyTerry2," Journal of Statistical Software, 2012.

[7] R. D. Luce, Individual choice behaviour: A theoretical analysis,
New York: Wiley, 1959.

[8] R. Plackett, "The Analysis of Permutations," Journal of the Royal
Statistical Society, pp. 193-202, 1975.

[9] R. K. S. Hankin, "Partial rank data with the hyper2 package:
likelihood functions for generalized Bradley-Terry," R Journal, 2017.

[10] L. Lundgren, R. Newton, T. Tran, M. Dunn, S. Nimphius and
J. Sheppard, "Analysis of Manoeuvres and Scoring in Competitive
Surfing," International Journal of Sports Science & Coachin, 2014.

[11] O. Farley, E. Raymond, J. Secomb, B. Ferrier and L. Lundgren,
"Scoring Analysis of the Men’s 2013 World Championship Tour of
Surfing," The International Journal of Aquatic Research and Education,
2015.

[12] I. McHale and A. Morton, "A Bradley-Terry type model for
forecasting tennis match results," International Journal of
Forecasting, pp. 619-630, 2011.

[13] H. Turner and D. Firth, "Bradley-Terry Models in R: The
BradleyTerry2 Package," Journal of Statistical Software, 2011.

[14] K. Butler and W. T. John , "The existence of maximum likelihood
estimates in the Bradley-Terry model and its extensions," 2004.

[15] R. K. S. Hankin, "A Generalization of the Dirichlet
Distribution," Journal of Statistical Software - Vol 33 - Issue 11,
2010.

[16] "Margaret River," World Surf League, 2019. [Online]. Available:
https://www.worldsurfleague.com/events/2019/mct/2913/margaret-river-pro.

[17] World Surf League, "Teahupo'o,"
August-September 2019. [Online]. Available:
https://www.worldsurfleague.com/events/2019/mct/2919/tahiti-pro-teahupoo.

[18] World Surf League, "Pipeline,"
December 2019. [Online]. Available:
https://www.worldsurfleague.com/events/2019/mct/2927/billabong-pipe-masters.

[19] "2019 Surfers," World Surf League, 2019. [Online]. Available:
https://www.worldsurfleague.com/athletes/tour/mct?year=2019.

[Tim's work ends, my work starts]

@@

@@

\subsection{Bradley-Terry and generalizations for rank
statistics}\label{bradley-terry-and-generalizations-for-rank-statistics}

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of [@johnson2020].  A slight generalization allows the
incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
did not finish, we would have

\begin{equation}\label{competitors_1_to_3_only_finished}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}
\end{equation}

(observe how this likelihood function, while informative about
$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
use a technique due to Hankin (2010, 2017) and introduce
fictional (reified) entities whose nonzero Bradley-Terry strength
helps certain competitors or sets of competitors under certain
conditions. The canonical example would be the home-ground advantage
in association football.  If players (teams) $1,2$ with strengths
$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
away wins for team $1$, and $c$ home wins and $d$ away wins for
team~$2$, then a suitable likelihood function would be

@@

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and effective coordination between
members of doubles tennis teams; we may use a similar device to
account for (e.g.) wet conditions in Formula 1.

# Literature Review and previous work

Most previous research into competitive surfing is limited to the
scoring of ariel manoeuvres compared with other manoeuvres in events
[10] and analysis in the difference in average wave scores for the top
10 and bottom 10 surfers using a One-way ANOVA method [11].

>>>>

@@

# Previous related work

@farley2013 analyse surfing results using one-way ANOVA techniques and
report comparisons between the top 10 and bottom 10 surfers in the
2013 World Championship Tour and conclude that the top-ranked athletes
are more consistent than the lower-ranked group. They report that "To
date, only a limited number of research studies have reported on the
results of competitive surfing" [p44], and the analysis presented here
is hoped to change that.

Other statistical research into competitive surfing appears to focus
on the scoring of aerial manoeuvres compared with other manoeuvres in
events.  @lundgren2014, for example, report that aerial manoeuvres
scored higher than other stunts but had a lower completion rate, a
theme we return to below.

@@

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of [@johnson2020].  We now use a technique due to Hankin
(2010, 2017) and introduce fictional (reified) entities whose nonzero
Bradley-Terry strength helps certain competitors or sets of
competitors under certain conditions.  The original example was the
home-ground advantage in football.  If players (teams) $1,2$ with
strengths $p_1,p_2$ compete, and if our observation were $a$ home wins
and $b$ away wins for team $1$, and $c$ home wins and $d$ away wins
for team~$2$, then a suitable likelihood function would be

@@

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and effective coordination between
members of doubles tennis teams.
